Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2015-01-20 draft_lm1.tex
added figure with the maps

\documentclass[10pt]{article}
% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}
% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
\usepackage{hyperref}
% line numbers
\usepackage{lineno}
% ligatures disabled
\usepackage{microtype}
\usepackage{todonotes}
\DisableLigatures[f]{encoding = *, family = * }
% rotating package for sideways tables
%\usepackage{rotating}
% If you wish to include algorithms, please use one of the packages below. Also, please see the algorithm section of our LaTeX guidelines (http://www.plosone.org/static/latexGuidelines) for important information about required formatting.
%\usepackage{algorithmic}
%\usepackage{algorithmicx}
% Use doublespacing - comment out for single spacing
%\usepackage{setspace} 
%\doublespacing
% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm
% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2009}
% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Estimating the Attack ratio of Dengue Epidemics without serotype 
information}
}
% Insert Author names, affiliations and corresponding author email.
\\
Flavio Code\c{c}o Coelho$^{1\ast}$, 
Luiz Max de Carvalho$^{2}$, 
\\
\bf{1} Escola de Matem\'atica Aplicada , Funda\c{c}\~ao Getulio Vargas (FGV), 
Rio de Janeiro -- RJ, Brazil.
\\
\bf{2} Programa de Computa\c{c}\~ao Cient\'ifica (PROCC), Funda\c{c}\~ao Oswaldo Cruz, Rio de Janeiro -- RJ, Brazil.
\\
$\ast$ E-mail: fccoelho@fgv.br
\end{flushleft}

@@

Dengue transmission is modulated by environmental conditions, of which 
temperature, due to its effects on the 
vector reproduction, stands out as a strong predictor of 
incidence~\cite{honorio_temporal_2009,wu_higher_2009}.
In places with sufficient temperature variation dengue is predominantly a 
summer disease.

The long term dynamics of dengue is also modulated by the alternation of virus 
types in circulation Demographics also plays a role in replenishing the 
population of susceptibles.

The Attack ratio of disease is a mesure of mobidity defined as the number of new
cases divided byt the population at risk.
For Dengue outbreaks and epidemics 
this can be difficult to calculated due to the lack of information about the 
population at risk. The population at risk in the case of dengue, is the number 
of susceptibles to the circulating virus type(s) befor a given epidemic.

The main source of information about Dengue dynamics are the incidence 
timeseries, which represent all the cases of Dengue reported by health 
professionals with no discrimination of virus types.
Only a very small sample 
of the cases are ever examined to determined the virus type.

@@

Various challenges exist to assess the attack rate of a dengue epidemic. 
One of 
them is that there are a large number of asymptomatic cases in any epidemic, 
which nevertheless acquire immunity.
It is estimated that for 
every case reported, 10-20 are not seen by health authorities (personnal 
communication).
Thirdly, under-reporting is a serious issue in Brazil. 
Duarte and Fran\c{c}a (2006)~\cite{duarte_data_2006}, estimated 
the sensitivity of dengue reporting for hospitalized patients in 
Belo-Horizonte, Brazil to be of 63\%, 
meaning that aproximately 37\% of the suspected dengue cases go unreported.  
Lastly, demography and migrations affect the number of susceptible in ways which 
are not easy to determine.

In this paper, we estimate the number of susceptibles before every major 
outbreak ($S_{0, year}$) in the last 18 years.
Due to the alternation of 
ciculating serotypes, we detected fluctuating levels of susceptibility the 
circulating virus(es).
To estimate the susceptibles we have we have fixed the 
tramissibility based on the $R_t$ series derived from data~\cite{nishiura}. 
Then, 
from the incidence series and the population at risk, we could then calculate 
the attack ratio for each epidemic.

@@

>>>>

\documentclass[10pt]{article}
\usepackage{todonotes}
\usepackage{subfigure}
% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}
% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
\usepackage{hyperref}
% line numbers
\usepackage{lineno}
% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }
% rotating package for sideways tables
%\usepackage{rotating}
% If you wish to include algorithms, please use one of the packages below. Also, please see the algorithm section of our LaTeX guidelines (http://www.plosone.org/static/latexGuidelines) for important information about required formatting.
%\usepackage{algorithmic}
%\usepackage{algorithmicx}
% Use doublespacing - comment out for single spacing
%\usepackage{setspace} 
%\doublespacing
% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm
% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2009}
% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Estimating the Attack ratio of Dengue Epidemics, under time-varying 
force of infection, from aggregated case notification data}
}
% Insert Author names, affiliations and corresponding author email.
\\
Flavio Code\c{c}o Coelho$^{1\ast}$, 
Luiz Max de Carvalho$^{2}$, 
\\
\bf{1} Escola de Matem\'atica Aplicada , Funda\c{c}\~ao Getulio Vargas (FGV), 
Rio de Janeiro -- RJ, Brazil.
\\
\bf{2} Programa de Computa\c{c}\~ao Cient\'ifica (PROCC), Funda\c{c}\~ao Oswaldo Cruz, Rio de Janeiro -- RJ, Brazil.
\\
$\ast$ E-mail: fccoelho@fgv.br
\end{flushleft}

@@

Dengue transmission is modulated by environmental conditions, of which 
temperature, due to its effects on the 
vector reproduction, stands out as a strong predictor of 
incidence~\cite{honorio_temporal_2009,wu_higher_2009}.
In places with sufficient temperature variation dengue is predominantly a 
summer disease. So it is fair to say that these environmental fluctuations play 
a key role in determining beginning and end of epidemic periods through its 
effects on the force of infection.

The long term dynamics of dengue is also modulated by the alternation of virus 
types in circulation. Demographics also plays a role in replenishing the 
population of susceptibles.

The Attack ratio of disease is a measure of morbidity defined as the number of 
new cases divided by the population at risk.
For Dengue outbreaks and epidemics this can be difficult to calculated due to 
the lack of information about the population at risk. The population at risk in 
the case of dengue, is the number of susceptibles to the circulating virus 
type(s) before a given epidemic.

The main source of information about Dengue dynamics are the incidence 
time series, which represent all the cases of Dengue reported by health 
professionals with no discrimination of virus types.
Only a very small sample 
of the cases are ever examined to determined the virus type.

@@

Various challenges exist to assess the attack rate of a dengue epidemic. 
One of 
them is that there are a large number of asymptomatic cases in any epidemic, 
which nevertheless acquire immunity.
It is estimated that for 
every case reported, 10-20 are not seen by health authorities (personnal 
communication).
Thirdly, under-reporting is a serious issue in Brazil. 
Duarte and Fran\c{c}a (2006)~\cite{duarte_data_2006}, estimated 
the sensitivity of dengue reporting for hospitalized patients in 
Belo-Horizonte, Brazil to be of 63\%, 
meaning that approximately 37\% of the suspected dengue cases go unreported.  
Lastly, demography and migrations affect the number of susceptible in ways which 
are not easy to determine.

In this paper, we estimate the number of susceptibles before every major 
outbreak ($S_{0, year}$) in the last 18 years.
Due to the alternation of 
circulating serotypes, we detected fluctuating levels of susceptibility the 
circulating virus(es).
To estimate the susceptibles we have we have fixed the 
transmissibility based on the $R_t$ series derived from data~\cite{nishiura}. 
Then, 
from the incidence series and the population at risk, we could then calculate 
the attack ratio for each epidemic.

@@

\begin{figure}
 \subfigure[2010]{\includegraphics[width=8cm, 
height=4.82cm]{./plots/heatmap2010.jpg}
% heatmap2010.jpg: 1075x589 pixel, 96dpi, 28.44x15.58 cm, bb=0 0 806 442
}
  \subfigure[2011]{\includegraphics[width=8cm]{./plots/heatmap2011.jpg}}
  \subfigure[2012]{\includegraphics[width=8cm]{./plots/heatmap2012.jpg}
% heatmap2010.jpg: 1075x589 pixel, 96dpi, 28.44x15.58 cm, bb=0 0 806 442
}
  \subfigure[2013]{\includegraphics[width=8cm]{./plots/heatmap2013.jpg}}
  \caption{{\bf Maps showing the incidence of dengue in the city of Rio de 
Janeiro from 2010 to 2013.} Circles indicate individual notified cases. A 
heatmap is overlayed on the maps showing absolute density of cases.}
  \label{fig:mapas}
\end{figure}
